Data Descriptive: Corresponding author journal publishing activities 2014 - 2018

### 2019-09-09 analysis/paper.Rmd
rmd reports render

---
title: "Data Descriptive: Corresponding author journal publishing activities 2014 - 2018"
date: "9/5/2019"
output: github_document
---

@@

The volume of corresponding author publications is a crucial indicator in the planning of open access funding programs and licensing models (Schimmer, Geschuhn, and Vogler 2015). But funders and national library consortia often lack access to such bibliometric data. Particularly, determining the share per publisher, journal and journal business model across countries is challenging. Here, by obtaining country affiliations from the Web of Science at the level of journals, and combining them with normalized publisher and open access status information, we created a comprehensive dataset about the productivity of corresponding authors for the period 2014-2018. This dataset is expected to have essential value for the quantitatve understanding of scholarly publishing. It will not only help to analyze the market share of publishing houses, but will also uncover the current state of the transition of subscription-based publishing models to fully open access.

Only a few openly available datasets focusing on the scholarly publishing industry exist. For example, Larivière et al. (2015), shared an aggregated dataset that contains the market share of publishers in terms of articles indexed in the Web of Science between 1973-2013. Using this data, the authors were able to reveal a growing market concentration among a few publishing houses. Palzenberger (2015) provided a dataset presenting the number and proportion of scholarly articles per country of affiliation including a breakdown per author role for the period 2004-2014. Published in 2015, this dataset, which was obtained from the Web of Science in-house database from the German Competence Center for Bibliometrics, formed the empirical basis for demonstrating that a large-scale transition from a subscription-based to open access business model is feasible. Since then, analyzing the productivity of corresponding authors in terms of articles published in dedicated journal portfolios has been perceived as a critical component of open access negotiations.

There have also been analytical applications for such datasets. Major bibliometric databases including the Web of Science, Scopus and Dimension have started to integrate open access status information. In these databases, Unpaywall data is often used to identify openly available full-texts at the article-level, while the Directory of Open Access Journals provides information about whether the journal is operating under a fully open access or toll-access business model. In its most recent edition, the Leiden Ranking presents open access indicators at the level of universities. In addition to university-specific data, the German Open Access Monitor, provides breakdowns per publisher and journal, as well as the inclusion of publications from non-university research institutions based in Germany

Here, we build an open dataset on the global productivity of corresponding authors. Drawing on Palzenberger's previous work, we not only created an updated dataset about the number and proportion of scholarly articles per country of affiliation from corresponding authors. The dataset also allows for a breakdown by journal, including information whether the journal is currently fully open access or not, and publishers. We constructed the dataset using the quality assured Web of Science in-house database from the German Competence Center for Bibliometrics. We enriched the data with publisher information from Crossref, a large DOI registration agency for scholarly work, and with open access journal information provided by the Biefeld GOLD OA list.

@@

Following Palzenberger, data were retrieved from the in-house Web of Science database maintained by the German Competence Center for Bibliometrics. The following data were obtained and aggregated:

@@

Our database retrieval reflected that it has become challenging to count corresponding authors straight, because the Web of Science has begun to keep track of more than one reprint author recently. While evaluating an efficient retrieval strategy, we found that the Web of Science did record more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018. Moreover, a percentage of 5% of reprint author was internationally co-located (see Figure 1). We therefore decided to use whole counting where every Web of Science record representing an original article or review was counted once per country of affiliation.

![](figure/country_wos.png)

@@

```{r}
publisher_df <- readr::read_csv("data/publisher_14_18.csv")
```

@@

```{r}
rp_df <- readr::read_csv("data/rp_jn_14_18.csv")
```

@@

```{r, results='asis'}
my_df <- readr::read_csv("data/publisher_14_18.csv")
tt_all <- my_df %>% 
  group_by(publisher) %>%
  summarise(n = sum(articles)) %>%
  mutate(prop = n /sum(n))  %>%
  mutate(publisher = ifelse(is.na(publisher), "Other", publisher)) %>%
  arrange(desc(n)) %>%
  mutate(publisher_group = fct_other(publisher, keep = publisher[1:11])) %>%
  ungroup() %>%
  group_by(publisher_group) %>%
  summarise(n = sum(n), prop = sum(prop)) %>%
  mutate(publisher_group = fct_reorder(publisher_group, n,)) %>%
  mutate(publisher_group = fct_relevel(publisher_group, "Other")) %>%
  arrange(desc(publisher_group))
knitr::kable(tt_all, caption = "Publisher Market Share 2014 - 2018 in terms of articles and reviews indexed in the Web of Science")
```

```{r}
tt_oa <- my_df %>%
  mutate(publisher_group = ifelse(publisher %in% tt_all$publisher_group, publisher, "Other")) %>%
  group_by(publisher_group, oa_journal) %>%
  summarise(n = sum(articles)) %>%
  mutate(oa_journal = ifelse(oa_journal == TRUE, "Open Access", "Toll Access")) %>%
  ungroup() %>%
  group_by(oa_journal) %>%
  mutate(prop = n / sum(n))
plot_df <- tt_all %>%
  mutate(oa_journal = "All") %>%
  bind_rows(tt_oa) %>%
  group_by(oa_journal) %>% 
  mutate(label = format(sum(n), big.mark   = ",")) %>% 
  mutate(label = paste0(oa_journal, " (", label, " Articles)")) %>%
  ungroup() %>%
  mutate(publisher_group = factor(publisher_group, levels = levels(tt_all$publisher_group)))
tableau_palette <- rev(c("#ccc94d","#82853b","#86b4a9","#39737c","#ffd94a","#b85a0d","#98d9e4","#3cb7cc","#ffb977","#ff7f0f","#acd98d","#32a251"))
p_bar <- ggplot(plot_df, aes(x="", y = prop, fill = publisher_group)) +
  geom_bar(width = 1, stat = "identity", position = position_stack(reverse = FALSE)) +
  facet_wrap(~label, ncol = 1) +
  coord_flip() +
  geom_text(aes(label = paste0(round(prop, 2) * 100, "")), position = position_stack(vjust = .5)) +
 scale_fill_manual("", values = c("grey80", tableau_palette[1:10]), guide = guide_legend(reverse = TRUE,  ncol= 4)) +
  scale_y_continuous(breaks = seq(0,1, by =0.1), labels = scales::percent_format(accuracy = 5L)) +
  labs(title = "Global Publisher Market Shares 2014 - 2018", x = "", y = "",
       subtitle = "Original Articles and Reviews indexed in the Web of Science\n\n",
       caption = "\nData Source:  Web of Science / Competence Centre for Bibliometrics for German Science.\nPublisher names were disambiguated using Crossref metadata. OA journals identified using the Bielefeld Gold OA list.\nFunding: Federal Ministry of Education and Research of Germany (BMBF)") +
  #  scale_y_continuous(breaks = seq(0,1, by =0.1), labels = scales::percent_format(accuracy = 5L)) +
  theme_minimal(base_family = "Roboto", base_size = 16) +
  theme(plot.margin = margin(30, 30, 30, 30)) +
  theme(panel.grid.minor = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(panel.grid.major.y = element_blank()) +
  theme(panel.border = element_blank()) +
  theme(legend.position="bottom",
        legend.justification = "center")
ggsave(
  filename = "figure/global_publisher_share.png",
  plot = p_bar,
  width = 14,
  height = 9,
  dpi = 300)
```

![](figure/global_publisher_share.png)

@@

![](figure/worldmap.png)

>>>>

---
title: "Data Descriptive: Corresponding author journal publishing activities 2014 - 2018"
date: "9/5/2019"
output: github_document
bibliography: pubs.bib
---

@@

The volume of corresponding author publications is a crucial indicator in the planning of open access funding programs and licensing models [@Schimmer_2015]. But funders and national library consortia often lack access to such bibliometric data. Particularly, determining the share per publisher, journal and journal business model across countries is challenging. Here, by obtaining country affiliations from the Web of Science at the level of journals, and combining them with normalized publisher and open access status information, we created a comprehensive dataset about the productivity of corresponding authors for the period 2014-2018. This dataset is expected to have essential value for the quantitatve understanding of scholarly publishing. It will not only help to analyze the market share of publishing houses, but will also uncover the current state of the transition of subscription-based publishing models to fully open access.

Only a few openly available datasets focusing on the scholarly publishing industry exist. For example, @haustein_2015 shared an aggregated dataset that contains the market share of publishers in terms of articles indexed in the Web of Science between 1973-2013. Using this data, the authors were able to reveal a growing market concentration among a few publishing houses [@Larivi_re_2015]. @Palzenberger_2015 provided a dataset presenting the number and proportion of scholarly articles per country of affiliation including a breakdown per author role for the period 2004-2014. Published in 2015, this dataset, which was obtained from the Web of Science in-house database from the German Competence Center for Bibliometrics, formed the empirical basis for demonstrating that a large-scale transition from a subscription-based to open access business model is feasible. Since then, analyzing the productivity of corresponding authors in terms of articles published in dedicated journal portfolios has been perceived as a critical component of open access negotiations.

There have also been analytical applications for such datasets. Major bibliometric databases including the Web of Science, Scopus and Dimension have started to integrate open access status information. In these databases, Unpaywall data is often used to identify openly available full-texts at the article-level, while the Directory of Open Access Journals provides information about whether the journal is operating under a fully open access or toll-access business model. In its most recent edition, the Leiden Ranking presents open access indicators at the level of universities [@leiden]. In addition to university-specific data, the German Open Access Monitor, provides breakdowns per publisher and journal, as well as the inclusion of publications from non-university research institutions based in Germany [@oa_monitor].

Here, we build an open dataset on the global productivity of corresponding authors. Drawing on Palzenberger's (2015) previous work, we not only created an updated dataset about the number and proportion of scholarly articles per country of affiliation from corresponding authors. The dataset also allows for a breakdown by journal, including information whether the journal is currently fully open access or not, and publishers. We constructed the dataset using the quality assured Web of Science in-house database from the German Competence Center for Bibliometrics. We enriched the data with publisher information from Crossref, a large DOI registration agency for scholarly work, and with open access journal information provided by the Biefeld GOLD OA list [@issn].

@@

Following @Palzenberger_2015, data were retrieved from the in-house Web of Science database maintained by the German Competence Center for Bibliometrics. The following data were obtained and aggregated:

@@

Our database retrieval reflected that it has become challenging to count corresponding authors straight, because the Web of Science has begun to keep track of more than one reprint author recently (see Figure 1). While evaluating an efficient retrieval strategy, we found that the Web of Science did record more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018. Moreover, a percentage of 5% of reprint author was internationally co-located (see Figure 2). We therefore decided to use whole counting where every Web of Science record representing an original article or review was counted once per country of affiliation.

![Figure 1](002_kb_rp_coverage_files/figure-gfm/ca_author_count-1.png)

![Figure 2](002_kb_rp_coverage_files/figure-gfm/ca_country_count-1.png)

@@

```{r}
publisher_df <- readr::read_csv("../data/publisher_14_18.csv")
```

@@

```{r}
rp_df <- readr::read_csv("../data/rp_jn_14_18.csv")
```

@@

```{r, results='asis'}
my_df <- readr::read_csv("../data/publisher_14_18.csv")
tt_all <- my_df %>% 
  group_by(publisher) %>%
  summarise(n = sum(articles)) %>%
  mutate(prop = n /sum(n))  %>%
  mutate(publisher = ifelse(is.na(publisher), "Other", publisher)) %>%
  arrange(desc(n)) %>%
  mutate(publisher_group = fct_other(publisher, keep = publisher[1:11])) %>%
  ungroup() %>%
  group_by(publisher_group) %>%
  summarise(n = sum(n), prop = sum(prop)) %>%
  mutate(publisher_group = fct_reorder(publisher_group, n,)) %>%
  mutate(publisher_group = fct_relevel(publisher_group, "Other")) %>%
  arrange(desc(publisher_group))
knitr::kable(tt_all, caption = "Publisher Market Share 2014 - 2018 in terms of articles and reviews indexed in the Web of Science")
```

```{r}
tt_oa <- my_df %>%
  mutate(publisher_group = ifelse(publisher %in% tt_all$publisher_group, publisher, "Other")) %>%
  group_by(publisher_group, oa_journal) %>%
  summarise(n = sum(articles)) %>%
  mutate(oa_journal = ifelse(oa_journal == TRUE, "Open Access", "Toll Access")) %>%
  ungroup() %>%
  group_by(oa_journal) %>%
  mutate(prop = n / sum(n))
plot_df <- tt_all %>%
  mutate(oa_journal = "All") %>%
  bind_rows(tt_oa) %>%
  group_by(oa_journal) %>% 
  mutate(label = format(sum(n), big.mark   = ",")) %>% 
  mutate(label = paste0(oa_journal, " (", label, " Articles)")) %>%
  ungroup() %>%
  mutate(publisher_group = factor(publisher_group, levels = levels(tt_all$publisher_group)))
tableau_palette <- rev(c("#ccc94d","#82853b","#86b4a9","#39737c","#ffd94a","#b85a0d","#98d9e4","#3cb7cc","#ffb977","#ff7f0f","#acd98d","#32a251"))
p_bar <- ggplot(plot_df, aes(x="", y = prop, fill = publisher_group)) +
  geom_bar(width = 1, stat = "identity", position = position_stack(reverse = FALSE)) +
  facet_wrap(~label, ncol = 1) +
  coord_flip() +
  geom_text(aes(label = paste0(round(prop, 2) * 100, "")), position = position_stack(vjust = .5)) +
 scale_fill_manual("", values = c("grey80", tableau_palette[1:10]), guide = guide_legend(reverse = TRUE,  ncol= 4)) +
  scale_y_continuous(breaks = seq(0,1, by =0.1), labels = scales::percent_format(accuracy = 5L)) +
  labs(title = "Global Publisher Market Shares 2014 - 2018", x = "", y = "",
       subtitle = "Original Articles and Reviews indexed in the Web of Science\n\n",
       caption = "\nData Source:  Web of Science / Competence Centre for Bibliometrics for German Science.\nPublisher names were disambiguated using Crossref metadata. OA journals identified using the Bielefeld Gold OA list.\nFunding: Federal Ministry of Education and Research of Germany (BMBF)") +
  #  scale_y_continuous(breaks = seq(0,1, by =0.1), labels = scales::percent_format(accuracy = 5L)) +
  theme_minimal(base_family = "Roboto", base_size = 16) +
  theme(plot.margin = margin(30, 30, 30, 30)) +
  theme(panel.grid.minor = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(panel.grid.major.y = element_blank()) +
  theme(panel.border = element_blank()) +
  theme(legend.position="bottom",
        legend.justification = "center")
dir.create("figure")
ggsave(
  filename = "figure/global_publisher_share.png",
  plot = p_bar,
  width = 14,
  height = 9,
  dpi = 300)
```

![Figure 3](figure/global_publisher_share.png)

@@

```{r}
my_df <- readr::read_csv("../data/rp_jn_14_18.csv")
#'  transform to eu
eu <-
  c(
    "AUT",
    "BEL",
    "BGR",
    "CYP",
    "CZE",
    "DEU",
    "DNK",
    "ESP",
    "EST",
    "FIN",
    "FRA",
    "GBR",
    "GRC",
    "HRV",
    "HUN",
    "IRL",
    "ITA",
    "LTU",
    "LUX",
    "LVA",
    "MLT",
    "NLD",
    "POL",
    "PRT",
    "ROU",
    "SVK",
    "SVN",
    "SWE"
  )
my_df %>%
  mutate(eu = case_when(country_code %in% eu ~ "EU28")) %>%
  mutate(countries = ifelse(eu %in% "EU28", "EU28", country_code)) -> df_to_plot
#' top 5 publisher closed
my_df %>%
  filter(oa_journal == FALSE) %>%
  group_by(publisher) %>%
  summarise(n = sum(articles)) %>%
  arrange(desc(n)) %>%
  head(7) %>%
  .$publisher -> top_6
#' data transformation to prepare plotting of corresponding authors' country of affilation
df_to_plot %>%
  mutate(fct_publisher = ifelse(publisher %in% top_6, publisher, "Others")) %>%
  mutate(fct_publisher = ifelse(is.na(fct_publisher), "Others", fct_publisher)) %>%
  group_by(fct_publisher, countries) %>%
  summarize(n = sum(articles)) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  mutate(fct_publisher = factor(fct_publisher, levels = c(top_6[!is.na(top_6)], "Others"))) %>%
  mutate(countries = fct_reorder(countries, n)) -> plot_df
#' add country lat lon
#' eu28 frankfurt main is the center
eu_df <-
  data_frame(
    latitude = 50.6393264,
    longitude = 8.40556572,
    name = "EU28",
    iso3c = "EU28"
  )
library(countrycode)
gc <- readr::read_tsv("../data/geocodes.csv") %>%
  # add iso code used in kb database
  mutate(iso3c = countrycode::countrycode(country, "iso2c", "iso3c")) %>%
  bind_rows(eu_df) %>%
  filter(!iso3c %in% eu)

gc %>%
  select(-country) %>%
  inner_join(plot_df, by = c("iso3c" = "countries")) %>%
  select(-prop) %>%
  spread(fct_publisher, n) %>%
  mutate(all_pubs = rowSums(.[5:11], na.rm = TRUE)) -> plot_wide
tmp <- plot_wide %>%
  arrange(desc(all_pubs)) %>%
  replace(., is.na(.), 0) 
#' plot
world <- map_data('world') %>%
  # without Antartica
  filter(region != "Antarctica")
p <- ggplot(world, aes(long, lat)) +
  geom_map(
    map = world,
    aes(map_id = region),
    color = "white",
    size = 0.1,
    fill  = "#b3b3b3a0"
  ) +
  coord_map() +
  scatterpie::geom_scatterpie(
    aes(
      x = longitude,
      y = latitude,
      r = sqrt(all_pubs / 5000)
    ),
    data = tmp,
    cols = c(levels(plot_df$fct_publisher)),
    color = NA,
    alpha = .7
  ) +
  geom_text(
    aes(
      x = longitude,
      y = latitude,
      group = iso3c,
      label = iso3c
    ),
    data = tmp,
    stat = "identity",
    size = sqrt(tmp$all_pubs / 50000)
  ) +
  scale_fill_manual(
    "Publisher",
    values = c(
      "Elsevier BV" = "#E69F00",
      "Springer Nature" = "#0072B2",
      "Wiley" = "#56B4E9",
      "Informa UK Limited" = "#009E73",
      "American Chemical Society (ACS)" = "#F0E442",
      "SAGE Publications" = "#D55E00",
      "Others" = "#fcfcfc"
    ),
    limits = levels(plot_df$fct_publisher)
  ) +
  coord_equal() +
  theme_minimal()  +
  labs(
    title = "Who did publish in toll-access journals between 2014 - 2018?",
    x = "",
    y = "",
    caption = "Data Sources:  Web of Science in-house database from the Competence Centre for Bibliometrics for German Science.\nPublisher names were disambiguated using Crossref metadata.Closed Access Journals identified using the Bielefeld GOLD OA list.\n
    Funding: Federal Ministry of Education and Research of Germany (BMBF), Project OAUNI (01PU17023A)",
    subtitle = "Circle size represents the number of corresponding authorships of original articles and reviews per country"
  ) +
  theme_minimal(base_family = "Roboto") +
  theme(plot.margin = margin(10, 30, 10, 30)) +
  theme(panel.grid.minor = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(panel.grid.major = element_blank()) +
  theme(panel.border = element_blank()) +
  theme(axis.ticks = element_blank(),
        axis.text = element_blank()) +
  theme(legend.position = "top",
        legend.justification = "right")

ggsave(
  "figure/map.png",
  p,
  dpi = 450,
  height = 6,
  width = 9
)

```

![Figure 4](figure/map.png)
